---
title: "Analysis of Illumina 450K HumanMethylation Data"
output: html_notebook
---


```{r load_libs, message=FALSE}
source("https://bioconductor.org/biocLite.R")
library(tidyverse)
library(reshape2)
library(minfi)
library(shinyMethyl)
library(limma)
library(statmod)
library(DMRcate)

# install.packages(c("RPMM", "WGCNA", "sqldf", "impute", "BMIQ", "dynamicTreeCut"))
library(RPMM)
library(WGCNA)
library(sqldf)
library(impute)
library(RnBeads)
library(dynamicTreeCut)
```


# Preprocessing
At this stage in the analysis it's assumed that there's a `Large GenomicRatioSet` named `norm_data_f`, and an annotation data frame named `annotation`. At the end of the preprocessing notebook, there's the option to save these objects so they can be loaded in here.  

```{r load_preprocessed_data}
load("Preprocessed_Data.RData")
```


# Hovarth's Methylation Clock

```{r load_essential_functions_and_ref_sets}
source("Hovarth/NORMALIZATION.R")
source("Hovarth/Horvath_MethyAge_Normalisation_Functions.R")

probeAnnotation21kdatMethUsed <- read_csv("../Scripts/Hovarth/probeAnnotation21kdatMethUsed.csv")
probeAnnotation27k            <- read_csv("../Scripts/Hovarth/datMiniAnnotation.csv")
datClock                      <- read_csv("../Scripts/Hovarth/AdditionalFile3.csv")

trafo      <- function(x, adult.age=20) { 
  x <- (x+1)/(1+adult.age)
  y <- ifelse(x<=1, log(x),x-1)
  y 
}

anti.trafo <- function(x, adult.age=20) { 
  ifelse(x<0, 
         (1+adult.age)*exp(x)-1, 
         (1+adult.age)*x+adult.age) 
}

asnumeric1 <- function(x) {
  as.numeric(as.character(x))
}

pheno_in          <- pData(norm_data_f) %>% as.data.frame
mVal              <- getM(norm_data_f)
design            <- model.matrix(~0 + SampleType + Source, data = pheno_in)
fitA              <- lmFit(mVal, design)
mAdj.fit          <- fitA$coefficients[,-c(1:length(unique(pheno_in$SampleType)))]
mAdj              <- as.matrix(mVal) - mAdj.fit %*% t(design[,-c(1:length(unique(pheno_in$SampleType)))])
betaAdj           <- ilogit2(mAdj)
dat0              <- as.data.frame(betaAdj)
dat0$CpG          <- as.character(rownames(dat0))
dat0              <- dat0[,c(ncol(dat0), 1:(ncol(dat0)-1))]
```


# Stage 1
Set up the analysis run with some basic parameters

```{r stage_1}
nSamples     <- dim(dat0)[[2]]-1
nProbes      <- dim(dat0)[[1]]
dat0[,1]     <- gsub(x=dat0 [,1],pattern="\"",replacement="")

match1       <- match(probeAnnotation21kdatMethUsed$Name, dat0[,1])
dat1         <- dat0[match1,]

set.seed(73)
```


# Stage 2
Normalise values using BMIQ so that intensities are comparable to the reference sets. This is really really really slow.

```{r stage_2}
meanMethBySample      <- as.numeric(apply(as.matrix(dat1[,-1]), 2, mean, na.rm=T))
minMethBySample       <- as.numeric(apply(as.matrix(dat1[,-1]), 2, min,  na.rm=T))
maxMethBySample       <- as.numeric(apply(as.matrix(dat1[,-1]), 2, max,  na.rm=T))

datMethUsed           <- t(dat1[,-1])
noMissingPerSample    <- apply(as.matrix(is.na(datMethUsed)), 1, sum)
gs_in                 <- probeAnnotation21kdatMethUsed$goldstandard2

datMethUsedNormalized <- BMIQcalibration(datM              = datMethUsed, 
                                         goldstandard.beta = gs_in, 
                                         plots             = F)
```


# Stage 3
Gather metrics and make predictions

```{r stage_3}
selectCpGsClock <- is.element(dimnames(datMethUsedNormalized)[[2]], 
                              as.character(datClock$CpGmarker[-1]))
datMethClock0   <- data.frame(datMethUsedNormalized[,selectCpGsClock])
datMethClock    <- data.frame(datMethClock0[as.character(datClock$CpGmarker[-1])])
predictedAge    <- as.numeric(anti.trafo(datClock$CoefficientTraining[1] + 
                                           as.matrix(datMethClock) %*% 
                                           as.numeric(datClock$CoefficientTraining[-1])))

Comment                      <- ifelse(predictedAge <0, 
                                       "Negative DNAm age.", 
                                       ifelse(predictedAge > 100, 
                                              "Old DNAm age.", 
                                              rep("",length(predictedAge))))
Comment[is.na(predictedAge)] <- "Age prediction was not possible. "
```



# Stage 4
Gather output together 

```{r stage_4}
datout                              <- data.frame(SampleID = colnames(dat1)[-1], 
                                                  DNAmAge  = predictedAge, 
                                                  Comment, 
                                                  noMissingPerSample, 
                                                  meanMethBySample, 
                                                  minMethBySample,
                                                  maxMethBySample)

datout                              <- cbind(datout, as.numeric(pData(lumi.norm.base)$Age))
datout                              <- cbind(datout, pData(lumi.norm.base)$SampleType)
colnames(datout)[8:9]               <- c("Chron_Age", "SampleType")
write.csv(datout, file="Methylation_Age.csv")

pData(lumi.norm.filtered)$MethAge   <- as.numeric(datout$DNAmAge)
pData(lumi.norm.filtered)$mcAgeDiff <- pData(lumi.norm.filtered)$MethAge - as.numeric(pData(lumi.norm.filtered)$Age)
```



```{r plot_cAge_vs_mAge}
gg <- ggplot(datout, aes(x=Chron_Age, y=DNAmAge, colour=SampleType)) +
            geom_point() + 
            theme_bw() + 
            scale_y_continuous(limits=c(0,100)) +
            scale_x_continuous(limits=c(0,100)) + 
            stat_smooth(method    = lm, 
                        fullrange = T,
                        se        = F) +
            geom_abline(intercept = 0) + 
            facet_grid(SampleType ~ .)
# png("MethAge_Vs_ChronAge.png",
#     width  = 12.53, 
#     height = 6.98, 
#     units  = "in", 
#     res    = 600) 
# print(gg)
# dev.off()
```









